\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates

### 2021-09-23 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
remove exp gomp weib models

---
author:
  - name: Sahir Rai Bhatnagar*
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon*
    affiliation: 'University of Manitoba'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: https://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James A. Hanley
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title: "\\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates"
abstract: |
  | In epidemiological studies of time-to-event data, a quantity of interest to the clinician is their patient's risk of an event. However, methods relying on time matching or risk-set sampling (including Cox regression) eliminate the baseline hazard from the  estimating function. The baseline hazard then needs to be estimated separately using a non-parametric approach. This leads to step-wise estimates of the cumulative incidence that are difficult to interpret. Using case-base sampling, Hanley & Miettinen (2009) explained how parametric hazard functions can be estimated using logistic regression. Their approach naturally leads to estimates of the cumulative incidence that are smooth-in-time.
  | In this paper, we present the \CRANpkg{casebase} R package, a comprehensive and flexible toolkit for parametric survival analysis. We describe how the case-base framework can be used in more complex settings: non-linear functions of time and non-proportional hazards, competing risks, and variable selection. Our package also includes an extensive array of visualization tools to complement the analysis. We illustrate all these features through three different case studies.
  | * SRB and MT contributed equally to this work.
preamble: >
  \usepackage{longtable}
  \usepackage{graphicx}
  \usepackage{tabularx}
  \usepackage{float}
  \usepackage{makecell}
  \usepackage{tabu}
output: rticles::rjournal_article 
bibliography: bhatnagar-turgeon-islam-hanley-saarela.bib 
fig_width: 12
fig_height: 9
fig_caption: true
editor_options: 
  chunk_output_type: console
---

@@

  - \code{glm}: This is the familiar logistic regression.
  - \code{glmnet}: This option allows for variable selection using the elastic-net [@zou2005regularization] penalty (cf. Case Study 3). This functionality is provided through the \pkg{glmnet} package [@friedman2010jss].
  - \code{gam}: This option provides support for *Generalized Additive Models* via the \CRANpkg{mgcv} package [@hastie1987generalized].
  - \code{gbm}: This option provides support for *Gradient Boosted Trees* via the \CRANpkg{gbm} package. This feature is still experimental.

@@

For our first case study, we return to the ERSPC study and investigate the differences in risk between the control and screening arms. We fit four models that differ in which functional form of time is used: 1) excluded from the linear predictor (Eq. \eqref{eq:exp}); 2) linear function (Eq. \eqref{eq:gomp}); 3) log function (Eq. \eqref{eq:weibull}); and 4) a smooth function using cubic B-splines. The models are fit using \code{fitSmoothHazard} with the familiar formula interface:

```{r, eval = eval_cs1, echo = -1}
set.seed(12)
fmla <- list(exponential = formula(DeadOfPrCa ~ ScrArm),
             gompertz = formula(DeadOfPrCa ~ Follow.Up.Time + ScrArm),
             weibull = formula(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm),
             splines = formula(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm))

fits <- lapply(fmla, function(form) {
  fitSmoothHazard(form, data = ERSPC, ratio = 100)
})
```

The output object from \code{fitSmoothHazard} inherits from the \code{singleEventCB} and \code{glm} classes. For this reason, we leveraged the \code{summary} method for \code{glm} objects to output a familiar summary of the results:

```{r, eval = TRUE, echo=eval_cs1}
summary(fits[["splines"]]) 
```

@@

```{r, eval=FALSE, echo=echo_plot_code}
new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
new_time <- seq(0, 14, by = 0.1)

risks <- lapply(fits, function(fit) {
  absoluteRisk(fit, time = new_time, newdata = new_data)
})
```

In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIFs from \pkg{casebase} on the Cox model CIF. The CIF estimates for the exponential model in panel (1) overestimate the cumulative incidence earlier on in time, and underestimate it later on. Based on our earlier discussion of the population-time plot, this poor fit for the exponential hazard was expected. We notice a better fit with increasing complexity of our model for time in Figure \ref{fig:erspc-cox-cif} (panels 2--4). As noted in Theoretical Details section, the usual asymptotic results hold for likelihood ratio tests built using case-base sampling models. Therefore, we can easily test the null hypothesis that the exponential model is just as good as the larger (in terms of number of parameters) spline model:

```{r erscp-compare, eval = eval_cs1, echo = FALSE}
anova(fits$exponential, fits$splines, test = "LRT")
```

The null hypothesis is rejected in favor of the spline model. Similarly, the AIC provides further evidence that the flexible function of time provides the best fit:

```{r erscp-aic, eval = eval_cs1, echo=FALSE}
c("Exp." = AIC(fits$exponential),
  "Gompertz" = AIC(fits$gompertz),
  "Weibull" = AIC(fits$weibull),
  "Splines" = AIC(fits$splines))
```

@@

```{r erspc-cox-cif, eval = eval_cs1, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="CIFs for control and screening groups in the ERSPC data. In each of the panels, we plot the CIF from the Cox model using \\code{survival::survfit} (solid line) and the CIF from the case-base sampling scheme (dashed line) with different functional forms of time. (1) The time variable is excluded (exponential). (2) Linear function of time (Gompertz). (3) The natural logarithm (Weibull). (4) Cubic B-spline expansion of time."}
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
myCols <- cbPalette[c(4,7)]

new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
new_time <- seq(0,14.5,0.1)

risks <- lapply(fits, function(i) {
  casebase::absoluteRisk(object = i, time = new_time, newdata = new_data, method = "mont")
})

layout(matrix(1:4, ncol = 2, byrow = FALSE), respect = FALSE)
par(mar = c(0, 4.1, 4.1, 0))
plot(survfit(cox_model, newdata = new_data),
     ylab = "",
     xlab = "",
     fun = "event",
     xmax = max(new_time),
     conf.int = FALSE, col = myCols, 
     lwd = 2,
     yscale = 100,
     xaxt = 'n')
legend("topleft", 
       legend = c("Control (Cox)","Control (Case-base)",
                  "Screening (Cox)", "Screening (Case-base)"), 
       col = rep(myCols, each = 2),
       lty = c(1, 2, 1, 2), 
       lwd = 2,
       cex = 1.1,
       bg = "gray90")
lines(risks$exponential[,1], risks$exponential[,2], col = myCols[1], lty = 2, lwd = 2)
lines(risks$exponential[,1], risks$exponential[,3], col = myCols[2], lty = 2, lwd = 2)
text(x = max(new_time)*0.97, y = 0.05/100, labels = "(1)", cex = 1.5)

par(mar = c(4.1, 4.1, 0, 0))
plot(survfit(cox_model, newdata = new_data),
     ylab = "", xlab = "",
     fun = "event",
     xmax = max(new_time),
     conf.int = FALSE, col = myCols, 
     lwd = 2,
     yscale = 100)
lines(risks$weibull[,1], risks$weibull[,2], col = myCols[1], lty = 2, lwd = 2)
lines(risks$weibull[,1], risks$weibull[,3], col = myCols[2], lty = 2, lwd = 2)  
text(x = max(new_time)*0.97, y = 0.05/100, labels = "(3)", cex = 1.5)
mtext("Cumulative incidence (%)", side = 2, line = -1.5, outer = TRUE)

par(mar = c(0, 0, 4.1, 3.9))
plot(survfit(cox_model, newdata = new_data),
     ylab = "",xlab = "",
     fun = "event",
      xaxt = 'n', yaxt = 'n',
     xmax = max(new_time),
     conf.int = FALSE, col = myCols, 
     lwd = 2,
     yscale = 100)
lines(risks$gompertz[,1], risks$gompertz[,2], col = myCols[1], lty = 2, lwd = 2)
lines(risks$gompertz[,1], risks$gompertz[,3], col = myCols[2], lty = 2, lwd = 2)
text(x = max(new_time)*0.97, y = 0.05/100, labels = "(2)", cex = 1.5)

par(mar = c(4.1, 0, 0, 3.9))
plot(survfit(cox_model, newdata = new_data),
     ylab = "",xlab = "",
     fun = "event",
     yaxt = 'n',
     xmax = max(new_time),
     conf.int = FALSE, col = myCols, 
     lwd = 2,
     yscale = 100)
lines(risks$splines[,1], risks$splines[,2], col = myCols[1], lty = 2, lwd = 2)
lines(risks$splines[,1], risks$splines[,3], col = myCols[2], lty = 2, lwd = 2)
text(x = max(new_time)*0.97, y = 0.05/100, labels = "(4)", cex = 1.5)
mtext("Years since randomization", side = 1, line = -1.5, outer = TRUE)
```

```{r erspc-95ci22, echo = FALSE, eval = eval_cs1}
CI_cox <- c(exp(coef(cox_model)), exp(confint(cox_model)))

aft_fits <- lapply(list("exp","gompertz","weibull"), function(i)
  flexsurv::flexsurvreg(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm,
                           data = ERSPC, dist = i)
  )

names(aft_fits) <- list("exponential","gompertz","weibull")

# function to exponentiate HR and CI and return a HR (lower, Upper) character
exp_est_ci <- function(model = c("exponential","gompertz","weibull","splines"),
                       fit, 
                       covariate = "ScrArmScreening group", 
                       type = c("casebase", "survreg")) {

  type <- match.arg(type)
  covariate <- match.arg(covariate)

  if (type == "casebase"){

    cis <- confint(fit[[model]])

    sprintf("%0.2f (%0.2f, %0.2f)",
            exp(coef(fit[[model]]))[[covariate]],
            exp(cis)[covariate,1],
            exp(cis)[covariate,2])
  } else {

    if (model == "weibull") {

      # The maximized log-likelihoods are the same between survreg and flexsurvreg, 
      # however the parameterisation is different: the first coefficient (Intercept) 
      # reported by survreg is log(mu), and survreg's "scale" is
      # dweibull's (thus flexsurvreg)'s 1 / shape
      res <- -fit[[model]]$res[covariate,] / (1 / fit[[model]]$res["shape",])

      sprintf("%0.2f (%0.2f, %0.2f)",
              exp(res[["est"]]),
              exp(res[["U95%"]]),
              exp(res[["L95%"]]))

    } else if (model == "splines") {

      c("--")

    } else {

      res <- fit[[model]]$res[covariate,]

      sprintf("%0.2f (%0.2f, %0.2f)",
              exp(res[["est"]]),
              exp(res[["L95%"]]),
              exp(res[["U95%"]]))
    }
  }
}

casebase_HR_CI <- lapply(list("exponential","gompertz","weibull","splines"), 
                         exp_est_ci, 
                         fit = fits,
                         covariate = "ScrArmScreening group", 
                         type = "casebase")

survreg_HR_CI <- lapply(list("exponential","gompertz","weibull","splines"), 
                         exp_est_ci, 
                         fit = aft_fits,
                         covariate = "ScrArmScreening group", 
                         type = "survreg")

ptable <- cbind(list("Exponential","Gompertz","Weibull","Splines"),
            do.call(rbind, casebase_HR_CI), do.call(rbind, survreg_HR_CI))
```

In Table \ref{tab:print-erspc-estimates}, we present a side-by-side comparison of the hazard ratios and confidence intervals estimated from \code{fitSmoothHazard} and the corresponding parametric model using \code{survival::survreg}, as well as the Cox model estimate. The hazard ratio estimates and confidence intervals are similar across all four models. This reinforces the idea that, under proportional hazards, we do not need to model the full hazard to obtain reliable estimates of the hazard ratio. Nevertheless, Figure \ref{fig:erspc-cox-cif} shows that different parametric models can still give rise to qualitatively different estimates for the CIF.

```{r print-erspc-estimates, eval = eval_cs1}
kable(ptable, "latex", booktabs = TRUE, align = c("l","c","c"),
      caption = c("Comparison of estimated hazard ratios and 95\\% confidence intervals for ERSPC data."),
      col.names = c("Model","casebase::fitSmoothHazard","survival::survreg")) %>%
  kable_styling() %>%
  column_spec(1, bold = TRUE) %>%
  collapse_rows(columns = 1, latex_hline = "major", valign = "middle") %>%
  footnote(general = sprintf("Cox model estimate: HR (95%% CI) = %0.2f (%0.2f, %0.2f)", CI_cox[1], CI_cox[2], CI_cox[3]),
           general_title = "") 
```

>>>>

---
author:
  - name: Sahir Rai Bhatnagar*
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon*
    affiliation: 'University of Manitoba'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: https://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James A. Hanley
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title: "\\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates"
abstract: |
  | In epidemiological studies of time-to-event data, a quantity of interest to the clinician is their patient's risk of an event. However, methods relying on time matching or risk-set sampling (including Cox regression) eliminate the baseline hazard from the  estimating function. The baseline hazard then needs to be estimated separately using a non-parametric approach. This leads to step-wise estimates of the cumulative incidence that are difficult to interpret. Using case-base sampling, Hanley & Miettinen (2009) explained how parametric hazard functions can be estimated using logistic regression. Their approach naturally leads to estimates of the cumulative incidence that are smooth-in-time.
  | In this paper, we present the \CRANpkg{casebase} R package, a comprehensive and flexible toolkit for parametric survival analysis. We describe how the case-base framework can be used in more complex settings: non-linear functions of time and non-proportional hazards, competing risks, and variable selection. Our package also includes an extensive array of visualization tools to complement the analysis. We illustrate all these features through three different case studies.
  | * SRB and MT contributed equally to this work.
preamble: >
  \usepackage{longtable}
  \usepackage{graphicx}
  \usepackage{tabularx}
  \usepackage{float}
  \usepackage{makecell}
  \usepackage{tabu}
output: 
  rticles::rjournal_article:
    number_sections: true
bibliography: bhatnagar-turgeon-islam-hanley-saarela.bib 
fig_width: 12
fig_height: 9
fig_caption: true
editor_options: 
  chunk_output_type: console
---

@@

  - \code{glm}: This is the familiar logistic regression.
  - \code{glmnet}: This option allows for variable selection using the elastic-net [@zou2005regularization] penalty (cf. Case Study 3). This functionality is provided through the \pkg{glmnet} package [@friedman2010jss].
  - \code{gam}: This option provides support for *Generalized Additive Models* via the \CRANpkg{mgcv} package [@hastie1987generalized].

@@

For our first case study, we return to the ERSPC study and investigate the differences in risk between the control and screening arms. We fit a flexible hazard by using a smooth function of time modeled with cubic B-splines. The model is fit using \code{fitSmoothHazard} with the familiar formula interface:

```{r, eval = eval_cs1, echo = -1}
set.seed(12)
library(splines)
fit <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
                       data = ERSPC, ratio = 100)
```

The output object from \code{fitSmoothHazard} inherits from the \code{singleEventCB} and \code{glm} classes. For this reason, we can leverage the \code{summary} method for \code{glm} objects to output a familiar summary of the results:

```{r, eval = TRUE, echo=eval_cs1}
summary(fit) 
```

@@

```{r, eval=eval_cs1, echo=echo_plot_code}
new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
new_time <- c(0, 1)

absoluteRisk(fit, time = new_time, newdata = new_data)
```

In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIF from \pkg{casebase} on the Cox model CIF. As noted in the Theoretical Details section, the usual asymptotic results hold for likelihood ratio tests built using case-base sampling models. Therefore, we can easily test the significance of the spline model:

```{r erscp-compare, eval = eval_cs1, echo = FALSE}
anova(fit, test = "LRT")
```

The null hypothesis is rejected in favor of the spline model. Similarly, to compare different models (e.g. time modeled linearly), we could compute Akaike's Information Criterion (AIC) for each model and compare them.

@@

```{r erspc-cox-cif, eval = eval_cs1, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="CIF for control and screening groups in the ERSPC data. We plot the CIF from the Cox model using \\code{survival::survfit} (solid line) and the CIF from case-base sampling (dashed line). We used a cubic B-spline expansion of time."}
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
myCols <- cbPalette[c(4,7)]

new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
new_time <- seq(0,14,0.1)
# Recompute with more time points
risk <- absoluteRisk(fit, time = new_time, newdata = new_data)

plot(survfit(cox_model, newdata = new_data),
     ylab = "Cumulative incidence (%)",
     xlab = "Years since randomization",
     fun = "event",
     xmax = max(new_time),
     conf.int = FALSE, col = myCols, 
     lwd = 2,
     yscale = 100,
     xaxt = 'n')
legend("topleft", 
       legend = c("Control (Cox)","Control (Case-base)",
                  "Screening (Cox)", "Screening (Case-base)"), 
       col = rep(myCols, each = 2),
       lty = c(1, 2, 1, 2), 
       lwd = 2,
       cex = 1.1,
       bg = "gray90")
lines(risk[,1], risk[,2], col = myCols[1], lty = 2, lwd = 2)
lines(risk[,1], risk[,3], col = myCols[2], lty = 2, lwd = 2)
```

### 2021-09-23 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
remove exp gomp weib models

\begin{itemize}
\tightlist
\item
  \code{glm}: This is the familiar logistic regression.
\item
  \code{glmnet}: This option allows for variable selection using the
  elastic-net \citep{zou2005regularization} penalty (cf.~Case Study 3).
  This functionality is provided through the \pkg{glmnet} package
  \citep{friedman2010jss}.
\item
  \code{gam}: This option provides support for \emph{Generalized
  Additive Models} via the \CRANpkg{mgcv} package
  \citep{hastie1987generalized}.
\item
  \code{gbm}: This option provides support for \emph{Gradient Boosted
  Trees} via the \CRANpkg{gbm} package. This feature is still
  experimental.
\end{itemize}

@@

For our first case study, we return to the ERSPC study and investigate
the differences in risk between the control and screening arms. We fit
four models that differ in which functional form of time is used: 1)
excluded from the linear predictor (Eq. \eqref{eq:exp}); 2) linear
function (Eq. \eqref{eq:gomp}); 3) log function (Eq.
\eqref{eq:weibull}); and 4) a smooth function using cubic B-splines. The
models are fit using \code{fitSmoothHazard} with the familiar formula
interface:

\begin{Schunk}
\begin{Sinput}
fmla <- list(exponential = formula(DeadOfPrCa ~ ScrArm),
             gompertz = formula(DeadOfPrCa ~ Follow.Up.Time + ScrArm),
             weibull = formula(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm),
             splines = formula(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm))

fits <- lapply(fmla, function(form) {
  fitSmoothHazard(form, data = ERSPC, ratio = 100)
})
\end{Sinput}
\end{Schunk}

The output object from \code{fitSmoothHazard} inherits from the
\code{singleEventCB} and \code{glm} classes. For this reason, we
leveraged the \code{summary} method for \code{glm} objects to output a
familiar summary of the results:

\begin{Schunk}
\begin{Sinput}
summary(fits[["splines"]]) 
\end{Sinput}
\begin{Soutput}
#> Fitting smooth hazards with case-base sampling
#> 
#> Sample size: 159893 
#> Number of events: 540 
#> Number of base moments: 54000 
#> ----
#> 
#> Call:
#> fitSmoothHazard(formula = form, data = ERSPC, ratio = 100)
#> 
#> Deviance Residuals: 
#>    Min      1Q  Median      3Q     Max  
#> -0.409  -0.172  -0.135  -0.084   3.808  
#> 
#> Coefficients:
#>                       Estimate Std. Error z value Pr(>|z|)    
#> (Intercept)           -10.2905     0.3185  -32.31  < 2e-16 ***
#> bs(Follow.Up.Time)1     4.3587     0.8008    5.44  5.2e-08 ***
#> bs(Follow.Up.Time)2     2.0890     0.4721    4.42  9.7e-06 ***
#> bs(Follow.Up.Time)3     4.6266     0.6910    6.70  2.1e-11 ***
#> ScrArmScreening group  -0.2354     0.0886   -2.66   0.0079 ** 
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#> (Dispersion parameter for binomial family taken to be 1)
#> 
#>     Null deviance: 6059.0  on 54539  degrees of freedom
#> Residual deviance: 5785.3  on 54535  degrees of freedom
#> AIC: 5795
#> 
#> Number of Fisher Scoring iterations: 9
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
new_time <- seq(0, 14, by = 0.1)

risks <- lapply(fits, function(fit) {
  absoluteRisk(fit, time = new_time, newdata = new_data)
})
\end{Sinput}
\end{Schunk}

In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIFs from
\pkg{casebase} on the Cox model CIF. The CIF estimates for the
exponential model in panel (1) overestimate the cumulative incidence
earlier on in time, and underestimate it later on. Based on our earlier
discussion of the population-time plot, this poor fit for the
exponential hazard was expected. We notice a better fit with increasing
complexity of our model for time in Figure \ref{fig:erspc-cox-cif}
(panels 2--4). As noted in Theoretical Details section, the usual
asymptotic results hold for likelihood ratio tests built using case-base
sampling models. Therefore, we can easily test the null hypothesis that
the exponential model is just as good as the larger (in terms of number
of parameters) spline model:

\begin{Schunk}
\begin{Soutput}
#> Analysis of Deviance Table
#> 
#> Model 1: DeadOfPrCa ~ ScrArm + offset(offset)
#> Model 2: DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm + offset(offset)
#>   Resid. Df Resid. Dev Df Deviance Pr(>Chi)    
#> 1     54538       6052                         
#> 2     54535       5785  3      267   <2e-16 ***
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{Soutput}
\end{Schunk}

The null hypothesis is rejected in favor of the spline model. Similarly,
the AIC provides further evidence that the flexible function of time
provides the best fit:

\begin{Schunk}
\begin{Soutput}
#>     Exp. Gompertz  Weibull  Splines 
#>     6056     5821     5807     5795
\end{Soutput}
\end{Schunk}

\begin{Schunk}
\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./erspc-cox-cif-1} \caption{CIFs for control and screening groups in the ERSPC data. In each of the panels, we plot the CIF from the Cox model using \code{survival::survfit} (solid line) and the CIF from the case-base sampling scheme (dashed line) with different functional forms of time. (1) The time variable is excluded (exponential). (2) Linear function of time (Gompertz). (3) The natural logarithm (Weibull). (4) Cubic B-spline expansion of time.}\label{fig:erspc-cox-cif}
\end{figure}
\end{Schunk}

In Table \ref{tab:print-erspc-estimates}, we present a side-by-side
comparison of the hazard ratios and confidence intervals estimated from
\code{fitSmoothHazard} and the corresponding parametric model using
\code{survival::survreg}, as well as the Cox model estimate. The hazard
ratio estimates and confidence intervals are similar across all four
models. This reinforces the idea that, under proportional hazards, we do
not need to model the full hazard to obtain reliable estimates of the
hazard ratio. Nevertheless, Figure \ref{fig:erspc-cox-cif} shows that
different parametric models can still give rise to qualitatively
different estimates for the CIF.

\caption{\label{tab:print-erspc-estimates}Comparison of estimated hazard ratios and 95\% confidence intervals for ERSPC data.}
\centering
\begin{tabular}[t]{>{}lcc}
\toprule
Model & casebase::fitSmoothHazard & survival::survreg\\
\midrule
\textbf{Exponential} & 0.79 (0.67, 0.94) & 0.81 (0.68, 0.96)\\
\cmidrule{1-3}
\textbf{Gompertz} & 0.80 (0.67, 0.95) & 0.80 (0.67, 0.95)\\
\cmidrule{1-3}
\textbf{Weibull} & 0.81 (0.68, 0.96) & 0.80 (0.65, 0.96)\\
\cmidrule{1-3}
\textbf{Splines} & 0.79 (0.66, 0.94) & --\\
\bottomrule
\multicolumn{3}{l}{\rule{0pt}{1em}Cox model estimate: HR (95\% CI) = 0.80 (0.67, 0.95)}\\
\end{tabular}
\end{table}

@@

\address{%
Sahir Rai Bhatnagar*\\
McGill University\\%
1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
%
\url{http://sahirbhatnagar.com/}%
%
\\\email{sahir.bhatnagar@mcgill.ca}
}

\address{%
Maxime Turgeon*\\
University of Manitoba\\%
186 Dysart Road Winnipeg, MB, Canada R3T 2N2\\
%
\url{https://maxturgeon.ca/}%
%
\\\email{max.turgeon@umanitoba.ca}
}

\address{%
Jesse Islam\\
McGill University\\%
1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
%
%
%
\\\email{jesse.islam@mail.mcgill.ca}
}

\address{%
James A. Hanley\\
McGill University\\%
1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
%
\url{http://www.medicine.mcgill.ca/epidemiology/hanley/}%
%
\\\email{james.hanley@mcgill.ca}
}

\address{%
Olli Saarela\\
University of Toronto\\%
Dalla Lana School of Public Health, 155 College Street, 6th floor,
Toronto, Ontario M5T 3M7, Canada\\
%
\url{http://individual.utoronto.ca/osaarela/}%
%
\\\email{olli.saarela@utoronto.ca}
}

>>>>

\begin{itemize}
\tightlist
\item
  \code{glm}: This is the familiar logistic regression.
\item
  \code{glmnet}: This option allows for variable selection using the
  elastic-net \citep{zou2005regularization} penalty (cf.~Case Study 3).
  This functionality is provided through the \pkg{glmnet} package
  \citep{friedman2010jss}.
\item
  \code{gam}: This option provides support for \emph{Generalized
  Additive Models} via the \CRANpkg{mgcv} package
  \citep{hastie1987generalized}.
\end{itemize}

@@

For our first case study, we return to the ERSPC study and investigate
the differences in risk between the control and screening arms. We fit a
flexible hazard by using a smooth function of time modeled with cubic
B-splines. The model is fit using \code{fitSmoothHazard} with the
familiar formula interface:

\begin{Schunk}
\begin{Sinput}
library(splines)
fit <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
                       data = ERSPC, ratio = 100)
\end{Sinput}
\end{Schunk}

The output object from \code{fitSmoothHazard} inherits from the
\code{singleEventCB} and \code{glm} classes. For this reason, we can
leverage the \code{summary} method for \code{glm} objects to output a
familiar summary of the results:

\begin{Schunk}
\begin{Sinput}
summary(fit) 
\end{Sinput}
\begin{Soutput}
#> Fitting smooth hazards with case-base sampling
#> 
#> Sample size: 159893 
#> Number of events: 540 
#> Number of base moments: 54000 
#> ----
#> 
#> Call:
#> fitSmoothHazard(formula = DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
#>     data = ERSPC, ratio = 100)
#> 
#> Deviance Residuals: 
#>    Min      1Q  Median      3Q     Max  
#> -0.403  -0.172  -0.135  -0.085   3.801  
#> 
#> Coefficients:
#>                       Estimate Std. Error z value Pr(>|z|)    
#> (Intercept)           -10.2533     0.3168  -32.36  < 2e-16 ***
#> bs(Follow.Up.Time)1     4.2544     0.7976    5.33  9.6e-08 ***
#> bs(Follow.Up.Time)2     2.1371     0.4720    4.53  5.9e-06 ***
#> bs(Follow.Up.Time)3     4.5339     0.6889    6.58  4.7e-11 ***
#> ScrArmScreening group  -0.2481     0.0886   -2.80   0.0051 ** 
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#> (Dispersion parameter for binomial family taken to be 1)
#> 
#>     Null deviance: 6059.0  on 54539  degrees of freedom
#> Residual deviance: 5787.6  on 54535  degrees of freedom
#> AIC: 5798
#> 
#> Number of Fisher Scoring iterations: 8
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
new_time <- c(0, 1)

absoluteRisk(fit, time = new_time, newdata = new_data)
\end{Sinput}
\begin{Soutput}
#>  time                
#>     0 0.0e+00 0.0e+00
#>     1 5.4e-05 4.2e-05
\end{Soutput}
\end{Schunk}

In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIF from
\pkg{casebase} on the Cox model CIF. As noted in the Theoretical Details
section, the usual asymptotic results hold for likelihood ratio tests
built using case-base sampling models. Therefore, we can easily test the
significance of the spline model:

\begin{Schunk}
\begin{Soutput}
#> Analysis of Deviance Table
#> 
#> Model: binomial, link: logit
#> 
#> Response: DeadOfPrCa
#> 
#> Terms added sequentially (first to last)
#> 
#> 
#>                    Df Deviance Resid. Df Resid. Dev Pr(>Chi)    
#> NULL                               54539       6059             
#> bs(Follow.Up.Time)  3      263     54536       5796   <2e-16 ***
#> ScrArm              1        8     54535       5788   0.0048 ** 
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{Soutput}
\end{Schunk}

The null hypothesis is rejected in favor of the spline model. Similarly,
to compare different models (e.g.~time modeled linearly), we could
compute Akaike's Information Criterion (AIC) for each model and compare
them.

\begin{Schunk}
\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./erspc-cox-cif-1} \caption{CIF for control and screening groups in the ERSPC data. We plot the CIF from the Cox model using \code{survival::survfit} (solid line) and the CIF from case-base sampling (dashed line). We used a cubic B-spline expansion of time.}\label{fig:erspc-cox-cif}
\end{figure}
\end{Schunk}

@@

\address{%
Sahir Rai Bhatnagar*\\
McGill University\\%
1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
%
\url{http://sahirbhatnagar.com/}\\%
%
\email{sahir.bhatnagar@mcgill.ca}%
}

\address{%
Maxime Turgeon*\\
University of Manitoba\\%
186 Dysart Road Winnipeg, MB, Canada R3T 2N2\\
%
\url{https://maxturgeon.ca/}\\%
%
\email{max.turgeon@umanitoba.ca}%
}

\address{%
Jesse Islam\\
McGill University\\%
1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
%
%
%
\email{jesse.islam@mail.mcgill.ca}%
}

\address{%
James A. Hanley\\
McGill University\\%
1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
%
\url{http://www.medicine.mcgill.ca/epidemiology/hanley/}\\%
%
\email{james.hanley@mcgill.ca}%
}

\address{%
Olli Saarela\\
University of Toronto\\%
Dalla Lana School of Public Health, 155 College Street, 6th floor,
Toronto, Ontario M5T 3M7, Canada\\
%
\url{http://individual.utoronto.ca/osaarela/}\\%
%
\email{olli.saarela@utoronto.ca}%
}
